Concentration of Maxima and Fundamental Limits in High-Dimensional Testing and Inference

### 2021-02-04 3.phase_transitions/3.global-tests.tex
accept changes in proof of Thm 3.1, Prop 5.1 and 5.3, Def. 5.3; other minor updates

 {\em (iv)} In the dense and faint signal setting of {\em (iii)}, the $L_1$ and $L_2$ statistics are both 
 sub-optimal. More precisely, they succeed in the sense of  \eqref{eq:support-recovery-success},  for $\underline{r}>\beta/2-{1}/{4}$, but fail in the 
 sense of \eqref{eq:support-recovery-failure} when $\overline{r}<\beta/2-{1}/{4}$.
\end{theorem}
%Theorem \ref{thm:detection-optimality} is proved in Section \ref{sec:supplement}.

@@

{\em Part {\rm (iv)}:}  We now turn to the $L_2$-norm statistic.
Recall a non-central chi-square random variable $\chi^2_k(\lambda)$ has mean $(k+\lambda)$ and variance $2(k+2\lambda)$.
Since the observations have distributions $\mathrm{N}(0,1)$ under the null and $\mathrm{N}(p^r,1)$ under the alternative, we have $x^2(i)\sim \chi^2_1(0)$ for $i\not\in S$ and $x^2(i)\sim \chi^2_1(p^{2r})$ for $i\in S$.
Therefore, the mean and variance of the (centered and scaled) $L_2$ statistics are
\begin{equation}
    \E\left[\frac{1}{\sqrt{p}}\sum_{i=1}^p \left(x(i)^2-1\right)\right] = 
    \begin{cases}
    0 & \text{under } H_0\\
    p^{1-\beta}p^{2r}p^{-1/2} = p^{1/2-\beta+2r} & \text{under } H_1,
    \end{cases}
\end{equation}
and 
\begin{equation}\label{e:detection-variance}
    \mathrm{Var}\left(\frac{1}{\sqrt{p}}\sum_{i=1}^p \left(x(i)^2-1\right)\right) = 
    \begin{cases}
    \frac{1}{p}2p = 2 & \text{under } H_0\\
    \frac{1}{p}\left(2p+4p^{1-\beta+2r}\right) = 2(1+ 2p^{2r-\beta}) & \text{under } H_1,
    \end{cases}
\end{equation}
respectively.
By the central limit theorem, we have
\begin{equation}
    {\color{red}\frac{1}{\sqrt{2p}}}\sum_{i=1}^p \left(x(i)^2-1\right) \implies \mathrm{N}(0,1),  
\end{equation}
under the null.    On the other hand, under the alternative, since $p^{2r-\beta}\to0$ for all $r<0$ and $\beta>0$, 
 the variance in \eqref{e:detection-variance} converges to $2$, as $p\to\infty$ and an application of the Lyapunov version of 
 the CLT, entails
 \begin{equation}
    {\color{red} \frac{1}{\sqrt{2p}}}\left(\sum_{i=1}^p \left(x(i)^2-1\right) - p^{1/2-\beta+2r}\right) \implies \mathrm{N}(0,1).
\end{equation}
Hence, perfect detection with the $L_2$-norm is possible if $p^{1/2-\beta+2r}$ diverges, i.e., $r>\beta/2-1/4$.
On the other hand, if $r<\beta/2-1/4$, the distributions of the (scaled) statistics merge under the null and the alternative. % \fbox{Hellinger distance is scale-invariant (show), detection is impossible}.

@@

By the central limit theorem, we have,
\begin{equation}
    \frac{1}{\sqrt{p}}\left(\sum_{i=1}^p |x(i)| - \sqrt{\frac{2}{\pi}}\right) \implies \mathrm{N}(0, 1-2/\pi)
\end{equation}
under the null.
On the other hand, when the alternative hypothesis holds, we have
\begin{align*}
    \E \left[ \frac{1}{\sqrt{p}} \left(\sum_{i=1}^p |x(i)| - \sqrt{\frac{2}{\pi}}\right) \right] 
    &= \frac{p^{1-\beta}}{\sqrt{p}}\left[\left(\sqrt{\frac{2}{\pi}}e^{-\mu^2/2} + \mu\left(1-2\Phi(-\mu)\right) \right) - \sqrt{\frac{2}{\pi}} \right] \\
    &= p^{1/2-\beta} \left[\sqrt{\frac{2}{\pi}}\left(e^{-p^{2r}/2} - 1 \right) + p^r\left(1-2\Phi(-\mu)\right) \right] \\
    &= p^{1/2-\beta} \left[\sqrt{\frac{2}{\pi}}\left(-p^{2r}/2 - O(p^{4r})\right) + p^r\left(\sqrt{\frac{2}{\pi}}p^r + O(p^{3r})\right) \right] \\
    &= p^{1/2-\beta} \sqrt{\frac{2}{\pi}} \left(p^{2r}/2 + O(p^{4r}) \right) \\
    &= p^{1/2-\beta+2r} \sqrt{{1}/{2\pi}} + O(p^{1/2-\beta+4r}),
\end{align*}
and 
\begin{align*}
    \mathrm{Var} \left(\frac{1}{\sqrt{p}} \left(\sum_{i=1}^p |x(i)| - \sqrt{\frac{2}{\pi}}\right) \right) 
    & = {\color{red}\frac{1}{p}}  (p-p^{1-\beta})(1-2/\pi) + {\color{red}\frac{1}{p}}   p^{1-\beta}\sigma_{Y}^2 \\
    & \to 1-2/\pi,
\end{align*}
by the boundedness of $\sigma_{Y}^2$ shown in \eqref{eq:bounded-variance-abs-Gaussian}.
Again, by the Lyapunov version of the central limit theorem, we conclude asymptotic normality of the centered and scaled $L_1$-norms under the alternative.
In an entirely analogous argument to the $L_2$-norm case, asymptotically perfect detection can be achieved if $p^{1/2-\beta+2r}$ diverges, i.e., $r>\beta/2-1/4$.
On the other hand, when $r<\beta/2-1/4$, the two hypotheses cannot be told apart by the $L_1$-norms since the distributions of the (scaled) statistics merge under the two hypotheses.
\end{proof}

>>>>

 {\em (iv)} In the dense and faint signal setting of {\em (iii)}, the $L_1$ and $L_2$ statistics are both 
 sub-optimal. More precisely, they succeed in the sense of  \eqref{eq:support-recovery-success} when $\underline{r}>\beta/2-{1}/{4}$, but fail in the 
 sense of \eqref{eq:support-recovery-failure} when $\overline{r}<\beta/2-{1}/{4}$.
\end{theorem}
%Theorem \ref{thm:detection-optimality} is proved in Section \ref{sec:supplement}.

@@

{\em Part {\rm (iv)}:}  We now turn to the $L_2$-norm statistic.
Recall a non-central chi-square random variable $\chi^2_k(\lambda)$ has mean $(k+\lambda)$ and variance $2(k+2\lambda)$.
Since the observations have distributions $\mathrm{N}(0,1)$ under the null and $\mathrm{N}(p^r,1)$ under the alternative, we have $x^2(i)\sim \chi^2_1(0)$ for $i\not\in S$ and $x^2(i)\sim \chi^2_1(p^{2r})$ for $i\in S$.
Therefore, the mean and variance of the (centered and scaled) $L_2$ statistics are
\begin{equation}
    \E\left[\frac{1}{\sqrt{p}}\sum_{i=1}^p \left(x(i)^2-1\right)\right] = 
    \begin{cases}
    0 & \text{under } H_0\\
    p^{1-\beta}p^{2r}p^{-1/2} = p^{1/2-\beta+2r} & \text{under } H_1,
    \end{cases}
\end{equation}
and 
\begin{equation}\label{e:detection-variance}
    \mathrm{Var}\left(\frac{1}{\sqrt{p}}\sum_{i=1}^p \left(x(i)^2-1\right)\right) = 
    \begin{cases}
    \frac{1}{p}2p = 2 & \text{under } H_0\\
    \frac{1}{p}\left(2p+4p^{1-\beta+2r}\right) = 2(1+ 2p^{2r-\beta}) & \text{under } H_1,
    \end{cases}
\end{equation}
respectively.
By the central limit theorem, we have
\begin{equation}
    \frac{1}{\sqrt{2p}}\sum_{i=1}^p \left(x(i)^2-1\right) \implies \mathrm{N}(0,1),  
\end{equation}
under the null.    On the other hand, under the alternative, since $p^{2r-\beta}\to0$ for all $r<0$ and $\beta>0$, 
 the variance in \eqref{e:detection-variance} converges to $2$, as $p\to\infty$ and an application of the Lyapunov version of 
 the CLT, entails
 \begin{equation}
    \frac{1}{\sqrt{2p}}\left(\sum_{i=1}^p \left(x(i)^2-1\right) - p^{1/2-\beta+2r}\right) \implies \mathrm{N}(0,1).
\end{equation}
Hence, perfect detection with the $L_2$-norm is possible if $p^{1/2-\beta+2r}$ diverges, i.e., $r>\beta/2-1/4$.
On the other hand, if $r<\beta/2-1/4$, the distributions of the (scaled) statistics merge under the null and the alternative. % \fbox{Hellinger distance is scale-invariant (show), detection is impossible}.

@@

By the central limit theorem, we have,
\begin{equation}
    \frac{1}{\sqrt{p}}\left(\sum_{i=1}^p |x(i)| - \sqrt{\frac{2}{\pi}}\right) \implies \mathrm{N}(0, 1-2/\pi)
\end{equation}
under the null.
On the other hand, when the alternative hypothesis holds, we have
\begin{align*}
    \E \left[ \frac{1}{\sqrt{p}} \left(\sum_{i=1}^p |x(i)| - \sqrt{\frac{2}{\pi}}\right) \right] 
    &= \frac{p^{1-\beta}}{\sqrt{p}}\left[\left(\sqrt{\frac{2}{\pi}}e^{-\mu^2/2} + \mu\left(1-2\Phi(-\mu)\right) \right) - \sqrt{\frac{2}{\pi}} \right] \\
    &= p^{1/2-\beta} \left[\sqrt{\frac{2}{\pi}}\left(e^{-p^{2r}/2} - 1 \right) + p^r\left(1-2\Phi(-\mu)\right) \right] \\
    &= p^{1/2-\beta} \left[\sqrt{\frac{2}{\pi}}\left(-p^{2r}/2 - O(p^{4r})\right) + p^r\left(\sqrt{\frac{2}{\pi}}p^r + O(p^{3r})\right) \right] \\
    &= p^{1/2-\beta} \sqrt{\frac{2}{\pi}} \left(p^{2r}/2 + O(p^{4r}) \right) \\
    &= p^{1/2-\beta+2r} \sqrt{{1}/{2\pi}} + O(p^{1/2-\beta+4r}),
\end{align*}
and 
\begin{align*}
    \mathrm{Var} \left(\frac{1}{\sqrt{p}} \left(\sum_{i=1}^p |x(i)| - \sqrt{\frac{2}{\pi}}\right) \right) 
    & = \frac{1}{p} (p-p^{1-\beta})(1-2/\pi) + \frac{1}{p} p^{1-\beta}\sigma_{Y}^2 \\
    & \to 1-2/\pi,
\end{align*}
by the boundedness of $\sigma_{Y}^2$ shown in \eqref{eq:bounded-variance-abs-Gaussian}.
Again, by the Lyapunov version of the central limit theorem, we conclude asymptotic normality of the centered and scaled $L_1$-norms under the alternative.
In an entirely analogous argument to the $L_2$-norm case, asymptotically perfect detection can be achieved if $p^{1/2-\beta+2r}$ diverges, i.e., $r>\beta/2-1/4$.
On the other hand, when $r<\beta/2-1/4$, the two hypotheses cannot be told apart by the $L_1$-norms since the distributions of the (scaled) statistics merge under the two hypotheses.
\end{proof}
